MLH1 Data Report

### 2019-11-12 doc/Main_Report_Results.rmd
updated/made results Rmd and report thing

```{r setup, echo=FALSE, include = FALSE}
library(knitr)
library(ggplot2)
#library(pwr)
library(plyr)
library(lattice)
library(dplyr)
setwd("C:/Users/alpeterson7/Documents/MLH1repo/")
#setwd("C:/Users/April/Desktop/MLH1repo")
#load(file="data/MLH1/MLH1_data_setup.RData")

#load main data file
load(file="data/MLH1/MLH1_data_setup_9.26.19.RData")#added batch16
#load(file="data/MLH1/MLH1_data_setup_8.29.19.RData") #

#load(file="data/MetaData.RData") #this was replacing most updated data file

@@

# Research Description

Measure nMLH1 foci per meiotic cell to estimate recombination rate for diverse strains of house mice and related rodents. Comparison the differences in recombination rates across sexes and genetic background to inform models of how meiotic recombination rates evolve.

## Discription of quantification process

Cell images are quantified in batches after file names are anonymized. The number of MLH1 foci, number of "achiasmate" and asynasped bivalents are quantified. A quality score is given; 1 to 5 (best to worst). All cell image file names are anonymized and moved into batches of a few hundred images.

```{r organizing.data, echo=FALSE}
#set the order of another column, based on another variable (so that when)

@@

This section shows the number of mice which haven't been stained or quantified yet.

```{r missing.mice, echo=FALSE, warning=FALSE}

setwd("C:/Users/alpeterson7/Documents/MLH1repo/")
setwd("~./MLH1repo/")
#other lists

Image_dirs <- list.files(path = "C:/Users/alpeterson7/Documents/Images")
#write this out so that I can access it

#remove the non-mouse folders
Dissection.File = read.csv("~./MLH1repo/data/Disections_53019.csv", header = TRUE)
Dissection.File <- Dissection.File[!(is.na(Dissection.File$mouse)|Dissection.File$mouse==""),]

#print the name with the date of the file somewhere
Dissection.list <- Dissection.File$mouse
Dissection.list <- as.character(Dissection.list)

#load metadata .. start decerning which mice are missing information... and which should be used
Meta.Data.File = read.csv("~./MLH1repo/data/ALP_MouseMetadata.csv", header = TRUE)

#remove file name, and original name .. this is just a mouse level DF
Meta.Data.File <- Meta.Data.File[!(is.na(Meta.Data.File$mouse)|Meta.Data.File$mouse==""),]
#make sure that the DOB is a date

meta.data.file2 = read.csv("~./MLH1repo/data/MouseMetaData61019.csv", header = TRUE)
meta.data.file2$DOB <- as.Date(meta.data.file2$DOB, format= "%m/%d/%y")

jl <- unique(Meta.Data.File$mouse[(Meta.Data.File$mouse %in% Dissection.File$mouse)])
BD.w.meta <- Meta.Data.File[(Meta.Data.File$mouse %in% Dissection.File$mouse), ]

#figure out a way to incorporate the staining... to get a better idea of which slides 
#to do next, #remove the pero mice?

#length(unique(MLH1_data$mouse) )

source("~./MLH1repo/src/CommonFunc_MLH1repo.R")
original_DF$fileName <- (original_DF$Original.Name)
original_DF <- add_mouse(original_DF)
#length(unique(original_DF$mouse) )
#131 mice have been quant total, 124 mice have passed into the final version

# make a list for the failed slides

#Dissection.list - Image/dirs
NonImaged.mice <- unique(Dissection.File$mouse[(Dissection.File$mouse %in% Image_dirs)])
#231 mice dissected which need images

#ImageFolder - MLH1.Quant
NonQuant.mice <- unique(Image_dirs[(Image_dirs %in% MLH1_data$mouse )])
#length(NonQuant.mice)
#123 mice that need to get into anon-batch

#make Dataframe to put into a table
mouse.number.DF <- data.frame( category = 
              c("Passed Quantified", "Total Quantified", "Non-Quantified", "Non-Imaged"),
              number.of.mice = c(length(unique(MLH1_data$mouse)),
                length(unique(original_DF$mouse) ),
                        length(NonQuant.mice),
                         length(NonImaged.mice)  )  )

kable(mouse.number.DF)
```

passed quantified - quantified mice (with enough cells)
total quantified -- all mice which have been (includes fail and passing mice)

non-quantified mice -- dissected mice / slides that haven't been quantified yet (all stained, but without image directory)

non-imaged -- mice without images, (might be stained and unstained)

This is the number of mice that are through different parts of the pathway or quantification process.

should I break down the categories into sex and strain?

@@

kable(table(AP_mouse_table$category), col.names = c("Category", "mice"), label = "Number of mice by category" )

@@

These are the tables for the number of cells and mice which have been quantified.
The 'Other' are F1 crosses.

## Repeatability of Human quantification

Well, we have to talk about the subjective nature that goes into scoring the MLH1 foci. Sources of error or variance across scoring come from background noise (false positives), faint foci (false negatives) and just human error.
A general anecdote is that if the quality of the cell staining / image is high, then the repeatability is near 100%. (using some of Beth's data).
Todo: track down, re-count of Beth's data,  cells that were counted first then in batch 0 (anonimized)

In the original_DF and MLH1 DF's there are ~400 and X respectively of cells (filenames) that are duplicated, or cells that have been counted in more than 1 quant batch

Datasets / sources of repeat measures
-chrom proportions
-multiple batch counts (later than batch0) *focus on this dataset for now. since its >100 and fairly easy to compile
-old and new (batch0)

The main take away is that higher quality scored cells have lower error in MLH1 counts.

> Does the amount of error differ across strains or sexes? (female might have a higher error rate)

> What sort of adjustments or tests should be made to results or analyses given the known error rate?

```{r quant.repeatability_rep1, echo=FALSE, include=TRUE}
#set up data for combining replicate counts

#1. replicates in DF, (17 mice, across 10 batches), focus on those from original for now

#isolate the duplicates (fileName) from main MLH1 DFs
n_occur <- data.frame(table(original_DF$fileName))
n_occur[n_occur$Freq > 1,]
original_DF_duplicates <- original_DF[original_DF$fileName %in% n_occur$Var1[n_occur$Freq > 1],]#2 is the max freq of duplicates,

#are there any batch, fileName duplicates?
original_DF_duplicates$unique.key <-  paste(original_DF_duplicates$fileName, original_DF_duplicates$Batch, sep = "_")
anyDuplicated(original_DF_duplicates$unique.key)#31 duplicated fileName_batch ... really 2?

n_occur <- data.frame(table(original_DF_duplicates$unique.key))
n_occur[n_occur$Freq > 1,]
double.trouble <- original_DF_duplicates[original_DF_duplicates$unique.key %in% n_occur$Var1[n_occur$Freq > 1],]
#2 fileNames from batch 0 appear 2x, remove them from DF

original_DF_duplicates <- original_DF_duplicates[ ! original_DF_duplicates$fileName %in% double.trouble$fileName, ]#removes 4 rows

#seperated duplicates into categories out in Excel (write to /results, moved final file to /data)
#This data frame was made ... with Jered's sugg
#write the duplicate file to csv/excel
#sort by file name
#need to remove '?' -- deal with the ? and X's
#write.table(original_DF_duplicates, "~./MLH1repo/results/orginal_DF_duplicates.csv", sep=",", row.names = FALSE)

original_DF_duplicates.notes = read.csv("~./MLH1repo/data/orginal_DF_duplicates.csv", header = TRUE)
#copies of fileNames can be distinguished with the 'separate.DF' column

#remove '?' that I stupidly left in
original_DF_duplicates.notes$nMLH1.foci = gsub("\\?", "", original_DF_duplicates.notes$nMLH1.foci)
original_DF_duplicates.notes$quality = gsub("\\?", "", original_DF_duplicates.notes$quality)
#these lines change col to chr

original_DF_duplicates.notes <- droplevels(original_DF_duplicates.notes)
#now there should be no more '?' going forward

#original_DF_duplicates.notes$quality <- as.factor(original_DF_duplicates.notes$quality)

#make lists of rows with "" or X's so I can remove all fileNames for those
#these lists just have na
quality.na.list <- original_DF_duplicates.notes$fileName[which(is.na(original_DF_duplicates.notes$quality) | original_DF_duplicates.notes$quality=="")] #16

MLH1.na.list <- original_DF_duplicates.notes$fileName[which(is.na(original_DF_duplicates.notes$nMLH1.foci) | original_DF_duplicates.notes$nMLH1.foci=="") ]#22

X.list <- original_DF_duplicates.notes$fileName[grepl("X", original_DF_duplicates.notes$X)]#40

#make DF which only has 
matches.only <- original_DF_duplicates.notes[ ! original_DF_duplicates.notes$fileName %in% quality.na.list, ]
matches.only <- matches.only[ ! matches.only$fileName %in% MLH1.na.list, ]
matches.only <- matches.only[ ! matches.only$fileName %in% X.list, ]

#str(matches.only)
#clean up match list (chr to num)
matches.only$nMLH1.foci <- as.numeric(matches.only$nMLH1.foci)
matches.only$quality <- as.numeric(matches.only$quality)

levels(matches.only$nMLH1.foci) #leves don't appear in num?.. move on, dwabt

#check that the 1, and 2 are equal number
length(matches.only$Original.Name[matches.only$separate.DF == 1]) #both 178

#i want to measure correleation of MLH1 across bins of quality (1,2,3,4,5)
#make new df with filename and MLH1.1 and mlh1.2

dups.here <- data.frame(  FN = unique(matches.only$Original.Name), MLH1.1 = matches.only$nMLH1.foci[matches.only$separate.DF == 1], MLH1.2 = matches.only$nMLH1.foci[matches.only$separate.DF == 2],
                        qual.1 = matches.only$quality[matches.only$separate.DF == 1] , 
                        qual.2 = matches.only$quality[matches.only$separate.DF == 2]
                          ) #OMG this worked!!!
dups.here$diff <- abs(dups.here$MLH1.1 - dups.here$MLH1.2) #the aveage difference is pretty small, but some have very high difference
#the high differences could potentially be mix ups

repeatability.table <- ddply(.data=dups.here, 
                 .(qual.1, qual.2),
                 summarize, 
                 ncells = length(unique(FN)),
                 plain_cor_MLH1 = cor(MLH1.1,MLH1.2)
)
#this table shows how many cells went changed into different qual score bins
#easiest patter, 1-1, 1-2 (high correlation in MLH1)
#5-5 lowest, (5-3 also low)
#4-5 has high corr? (direction shouldn't matter)
#4-2 has negative corr?

#of the different quality permutation bins, 37% don't change quality bin, 44% of cells change by 1 (either direction)

no.change = c(14,13,17,13,10)#67
change.1 = c(10,7,8,20,9,14,4,7)#79

repeatability.table.qual1 <- ddply(.data=dups.here, 
                 .(qual.1),
                 summarize, 
                 ncells = length(unique(FN)),
                 plain_cor_MLH1 = cor(MLH1.1,MLH1.2),
                 dif = mean(MLH1.1 - MLH1.2)
)

repeatability.table.qual2 <- ddply(.data=dups.here, 
                 .(qual.2),
                 summarize, 
                 ncells = length(unique(FN)),
                 plain_cor_MLH1 = cor(MLH1.1,MLH1.2)
)
#These tables show that qual1 cells have the highest corrleation of MLH1
dups.here.plain_cor_qual <- cor(dups.here$qual.1, dups.here$qual.2)

#i don't know how to compare quality in more

#nMLHJ1, quality are factors
original_DF_duplicates.notes$quality2 = gsub("\\?", "", original_DF_duplicates.notes$quality)

original_DF_duplicates.notes <- droplevels(original_DF_duplicates.notes)

#factor > chr > number  (this col is now chr is can go to num)
#original_DF_duplicates.notes$MLH1.foci2 <- as.numeric(original_DF_duplicates.notes$MLH1.foci2)
#original_DF_duplicates.notes$quality2 <- as.numeric(original_DF_duplicates.notes$quality2)

#one <- original_DF_duplicates.notes[original_DF_duplicates.notes$separate.DF == 1,]
#two <- original_DF_duplicates.notes[original_DF_duplicates.notes$separate.DF == 2,]
#rename this
#plain_cor_MLH1 <- cor(one$MLH1.foci2,two$MLH1.foci2,use = "complete.obs")
#.857
#plain_cor_qual <- cor(one$quality2,two$quality2,use = "complete.obs")
#.63

#What's the relationship of MLH1 cor with quality?
#what the rpeat of quality score?
#what are the correlations across quality score bins?

#issues that complicate the plain calculation
#some duplicates were 'x'ed and some weren't -- this causes different numbers
#connected with above, some duplicates have mlh1 scores, other don't -- this kinda messes up correlation

original_DF_duplicates.notes <- original_DF_duplicates.notes[ !grepl("X", original_DF_duplicates.notes$X) , ]
original_DF_duplicates.notes <- original_DF_duplicates.notes[!(is.na(original_DF_duplicates.notes$nMLH1.foci) | original_DF_duplicates.notes$nMLH1.foci==""), ]
#this causes unequal lengths

original_DF_duplicates.notes <- droplevels(original_DF_duplicates.notes)
original_DF_duplicates.notes$nMLH1.foci <- as.character(original_DF_duplicates.notes$nMLH1.foci)

original_DF_duplicates.notes$nMLH1.foci <- as.numeric(original_DF_duplicates.notes$nMLH1.foci)

#correlationsnMLH1 
#one <- original_DF_duplicates.notes[original_DF_duplicates.notes$separate.DF == 1,]
#two <- original_DF_duplicates.notes[original_DF_duplicates.notes$separate.DF == 2,]
#plain_cor <- cor(one$nMLH1.foci,two$nMLH1.foci)
#old.master.MLH1 <- original_DF[ ! original_DF$fileName %in% ll$fileName, ]

#check that all batches are distinguishable across duplicate fileNames
#make unique key fileName_batch -- then check for duplicate of that 
#original_DF$Batch <- factor(original_DF$Batch, ordered = TRUE, levels = c("Batch0", "Batch1", "Batch2", "Batch3", "Batch4","Batch5","Batch6",  "Batch7", "Batch8","Batch9","Batch10", "Batch11","Batch12", "Batch13", "Batch14","Batch15"))
#after doing this, original_DF$Batch[2000] > original_DF$Batch[10] will be evaluated!

@@

```{r repeatability_rep2, echo=FALSE, include=FALSE, eval=FALSE}
#2. old vs new, (remove duplicates and merge batch0 with old_master_mlh1)
#load, clean up and mat one of the master.MLH1count sheets
#master
old.master.MLH1 = read.csv("~./MLH1repo/data/old_master_MLH1_nov15.csv", header = TRUE)
#remove all with 'X'
#create a real fileName col, add rev.tif, 
source("~./MLH1repo/src/CommonFunc_MLH1repo.R")

old.master.MLH1 <- add_mouse(old.master.MLH1)
old.master.MLH1 <- add_strain(old.master.MLH1)
old.master.MLH1 <- add_sex(old.master.MLH1)
old.master.MLH1 <- add_category(old.master.MLH1)
old.master.MLH1$fileName2 <- old.master.MLH1$fileName

test.old.master <- rbind(head(old.master.MLH1), tail(old.master.MLH1) )
test.old.master$fileName <- as.character(test.old.master$fileName)

test.old.master$stain <- as.character(test.old.master$stain)

str(test.old.master)

#paste together file name, requires mouse, strain
#this might also need some character / factor stuff ... filename and stain MUST be charaecter!
source("~./MLH1repo/src/CommonFunc_MLH1repo.R")
#wrote add_fullFileName to make the file names for the old excel sheets

old.master.MLH1 <- add_fullFileName(old.master.MLH1)
old.master.MLH1$fileName <- old.master.MLH1$file.name_full

#merge and match to MLH1

#need to test for duplicates before mergeing?
n_occur <- data.frame(table(old.master.MLH1$file.name_full))
n_occur[n_occur$Freq > 1,]
ll <- old.master.MLH1[old.master.MLH1$file.name_full %in% n_occur$Var1[n_occur$Freq > 1],]
old.master.MLH1 <- old.master.MLH1[ ! old.master.MLH1$fileName %in% ll$fileName, ]

n_occur <- data.frame(table(original_DF$fileName))
n_occur[n_occur$Freq > 1,]
original_DF_duplicates <- original_DF[original_DF$fileName %in% n_occur$Var1[n_occur$Freq > 1],]
table(original_DF_duplicates$mouse)#17 mice

old.master.MLH1 <- original_DF[ ! original_DF$fileName %in% ll$fileName, ]

#old.master.MLH1 <- original_DF[ ! original_DF$fileName %in% ll$fileName, ]

#there are duplicates in original_DF and MLH1

uutesting2 <- merge(old.master.MLH1, original_DF, by.x = fileName, by.y = Original.Name)
#match old.full.filname with MLH1$fileName
```

```{r repeatability_rep3, echo=FALSE, include=FALSE}
chrm.class.DF = read.csv("~./MLH1repo/data/ChrmClassProportions.csv", header = TRUE)
source("~./MLH1repo/src/CommonFunc_MLH1repo.R")
#what are the target columns for this excel sheet?
#espically for getting repeat data
#MLH1.dif  same_count  quality

chrm.class.DF$fileName <- (chrm.class.DF$Original.Name)
chrm.class.DF <- add_mouse(chrm.class.DF)
chrm.class.DF <- add_strain(chrm.class.DF)
chrm.class.DF <- add_sex(chrm.class.DF)
chrm.class.DF <- add_category(chrm.class.DF)

#assessing error,
#number for same_count == TRUE
#length(chrm.class.DF$Original.Name[chrm.class.DF$same_count == "TRUE"])
#length(chrm.class.DF$Original.Name)

#cor of first count and calq count
#cor(chrm.class.DF$nMLH1.foci, chrm.class.DF$MLH1.sum.check)
#.87

#the true false might have to be changed first
#chrm.class.DF$same_count_num <- as.numeric(chrm.class.DF$same_count)

#regression of quality and MLH1.dif
#lmreg <- lm(abs(MLH1.dif) ~ quality + nMLH1.foci, data= chrm.class.DF)
#summary(lmreg)
#according to this linear regression, quality is not a significant

#make regression for just same counts
#I tried making a  binomial regression model, 
#but quality is bad predictor, only 5 levels)
#lmreg2 <- glm(same_count ~ quality, data= chrm.class.DF, family=binomial(link="logit"))
#summary(lmreg2)
#to plot the model / bimodal data
#fit = glm(same_count_num ~ as.numeric(quality), data=chrm.class.DF, family=binomial)
#newdat <- data.frame(quality=seq(min(chrm.class.DF$quality), max(chrm.class.DF$quality),len=102))
#newdat$vs = predict(fit, data=newdat, type="response")
#the prediction is not going 1 or 0 -- so doesn't fit logistic
#plot(vs ~ quality, data=newdat, col="red4")
#lines(vs ~ quality, newdat, col="green4", lwd=2)

hist_error_plot <- ggplot(data=chrm.class.DF, aes(MLH1.dif))+geom_histogram()+ggtitle("Distribution of foci count differences across 2 count versions")

#
quality_error_plot <- ggplot(data=chrm.class.DF, aes(x=quality, y=abs(MLH1.dif)  ))+geom_jitter()+ggtitle("Repeatability by quality score")+ylab("difference in quality score")+xlab("quality score")

error_count_plot <- ggplot(data=chrm.class.DF, aes(y=nMLH1.foci, x=abs(MLH1.dif)  ))+geom_jitter()+ggtitle("Error by MLH1 number")+xlab("difference in two counts")+ylab("MLH1 count")

#in the chrm proportions data is another set for eastimating error
```

Data use for this repeatabiity estimate is from smaller dataset where the proportions of chromosome/bivalent classes (0CO bivalents ect.) were quantified. (The focus of one count was cell total, while the other focus was classifying bivalents).

This data has `r length(chrm.class.DF$Original.Name)` cells across these categories

`r kable(table(droplevels(chrm.class.DF$category)))`

More cells are planned to be quantified.

```{r show.hist_error_plot, echo=FALSE,  fig.width=10, fig.height=5, warning=FALSE}
hist_error_plot
```

From this histogram we see most of the cells in this dataset have a difference of 2 in the MLH1 count scores (`r length(chrm.class.DF$Original.Name[( abs( chrm.class.DF$MLH1.dif) <= 2)]) / length(chrm.class.DF$Original.Name)`).

```{r show.quality_error_plot, echo=FALSE,  fig.width=10, fig.height=5}
quality_error_plot
```

There is a positive relationship with the absolute difference and quality score (where 1 is best and 5 is worst quality) (i.e. poorer quality cells are more likely to have more error acorss multiple counts). The proportion of the cells which had the same count across the counting instances is `r length(chrm.class.DF$Original.Name[chrm.class.DF$same_count == "TRUE"]) / length(chrm.class.DF$Original.Name)`. The correlation between the two versions of MLH1 counts for the same cell is `r cor(chrm.class.DF$nMLH1.foci, chrm.class.DF$MLH1.sum.check)`.

```{r show.error_count_plot, echo=FALSE,  fig.width=10, fig.height=5}
error_count_plot
```
There doesn't seem to be a bias across cells with more or less foci.

Modeling this relationship; (still in progress)
I tried doing a logistic regression for same_count = true or false depending on the quality score. I am having trouble making a plot the makes sense to confirm the binomial nature of the model.

## Effects on the dataset

error associated with 'quality of cels'

Main metric is mean CO count per mouse. Run tests to check if there are effects for the following effects.

*age effects* (male, maternal age)

*litter effect*

*mouse room effect* (Charmony, MSC)

*quantification repeatability*

*spread batch and staining effects*

*power analysis on number of cells*

*time of year effect (on spread)*

*dominant effect* (F1s)

These aspects will be analyzed with a linear models on subsets of data (within category) and visualized with plots.

### Quality effect

There is a positive correlation with MLH1 number and quality
To test this effect, compare all of the counts from quality-1 cells across strains. check that the same patterns of means persist compared to the total-cell data.

```{r quality.effect, echo=FALSE}

setwd("C:/Users/alpeterson7/Documents/MLH1repo/doc/")
#compare all quality 1 cells across strains
#check that the means (or directions of differences is maintained)

Q12_MLH1_DF <- MLH1_data[MLH1_data$quality <= 2,]

#ddply
Q12_MLH1_means <- ddply(.data=Q12_MLH1_DF, 
                 .(category),
                 summarize, 
                 ncells = length(unique(Original.Name)),
                 nmice = length(unique(mouse)),
                 mean.MLH1 = mean(adj_nMLH1.foci),
                 var.MLH1 = var(adj_nMLH1.foci)
)

tot.DF_MLH1_means <- ddply(.data=MLH1_data, 
                 .(category),
                 summarize, 
                 ncells = length(unique(Original.Name)),
                 nmice = length(unique(mouse)),
                 mean.MLH1 = mean(adj_nMLH1.foci),
                 var.MLH1 = var(adj_nMLH1.foci)
)

mergy <- merge(Q12_MLH1_means, tot.DF_MLH1_means, by='category', all = F)
colnames(mergy) <- c("category","Q12ncells","Q12mice","Q12_MLH1_means","Q12MLH1var",
                     "Totncells","Totnmice","TotMLH1mean","TotMLH1var")

mergy$percent.diff <- (mergy$Q12_MLH1_means / mergy$TotMLH1mean)

Q12toTotal <- ggplot(mergy, aes(y=Q12_MLH1_means, x=TotMLH1mean))+geom_point(aes(size = Q12ncells) )+geom_abline(slope = 1, intercept = 0)+ggtitle("Quality 1 cells MLH1 mean to total MLH1 mean")+ylim(c(21,32))+xlim(c(21,32))

Q12toTotal

#check that the directions are the same
#make tables for the 
#is the amount of 'polymorphism' the same

Q12_MLH1_means_males <- ddply(.data=Q12_MLH1_DF[Q12_MLH1_DF$sex == "male",], 
                 .(category,subsp),
                 summarize, 
                 ncells = length(unique(Original.Name)),
                 nmice = length(unique(mouse)),
                 mean.MLH1 = mean(adj_nMLH1.foci)
)

tot.DF_MLH1_means_males <- ddply(.data=MLH1_data[MLH1_data$sex == "male",], 
                 .(category,subsp),
                 summarize, 
                 ncells = length(unique(Original.Name)),
                 nmice = length(unique(mouse)),
                 mean.MLH1 = mean(adj_nMLH1.foci)
)

kable(Q12_MLH1_means_males, caption = "High quality cells")

kable(tot.DF_MLH1_means_males, caption = "Total Cell")

@@

All the quality 1 cells produce a higher mean. The biggest difference in means is HMI male, LEW female and SPIC female means have ~10% differences.

When quality 1 to 2 is compared, the strains were closer to the line, but the means are still higher in the higher quality cells.
with 1-2 most all of the difference are below 10% (except SPRET female).

MOLF males have quite a big difference.

There seem to be the similar amount of polymorphism in the Dom and more conserved in Dom.

Ranges of category means across subspecies.

| Type | male Dom | female Dom | | male Musc low | male Musc high | Musc female | | male Outgroup | female outgroup |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| Total cells | 22-24 |  24-27 |  | 23-31 | 25-31 | 26-29 | | 25-26 |27 |
| High quality cells | 23-25 | 25-29 | | 22-25 | 22-27 | 25 | | 26-27 | 29-31 |
(ASTmale in Musc high, has a 27 mean which might be an outlier)
I think the ranges across the groups is generally consistant.

(try looking at <5 quality scores?)

```{r high.qual, fig.caption="boxplots of MLH1 distributions", fig.height=5, fig.width=5, echo=FALSE, warnings=FALSE, include=FALSE}

#MLH1_data$quality <- as.numeric(MLH1_data$quality) #this needs to be run for the subsetting to work
highest_qual_df <-  subset(MLH1_data, as.numeric(MLH1_data$quality) <= 2 )

#are there no KAZ cells of high quality?
qq <- ggplot(highest_qual_df, aes(x = mouse, y = adj_nMLH1.foci)) + geom_boxplot(data = highest_qual_df, aes(fill = factor(strain) ) ) +  ggtitle("Cells with quality scores of 2 or 1") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + scale_fill_manual(values=c("#56B4E9", "cadetblue", "lightblue",  "coral1", "#E69F00", "red","yellowgreen", "green","purple", "black", "grey", 
             "grey2","grey3","grey1","grey","grey","grey"))
high_qual_boxplot <- qq + facet_wrap(~ sex, scales="free") #this is breaking the kniting... 
#high_qual_boxplot
#these boxplots look bad
```

> *How do the distributions change across different quality scores?*

Higher quality scored cells have more MLH1 foci. (some of the ranges get smaller with higher quality cells)

> *How do the distributions change across mice with a good number of cells?*

(was I referring to normale distributions?)
means all shift up, what about variance, range, lowest cut off

> *Do the directions of the differences across groups change due to quality scores?*

(polymorphism vs divergence patterns)

### Age Effect

Some papers have documented age effects for male genome-wide recombination rate [@zelazowski2017age].

The initial analysis uses plots. This should be double checked with a general linear model as well.

```{r age.hists, echo=FALSE, eval=FALSE,  warning=FALSE}
source("~./MLH1repo/src/CommonFunc_MLH1repo.R")
meta.data$DOB <- as.Date(meta.data$DOB, format= "%m/%d/%Y")#%Y 2000, %y '18,

#make Mouse meta data work .. rename these
MouseMetaData.wrkin <- meta.data[,c(1,4)]
colnames(MouseMetaData.wrkin) <- c("mouse", "DOB")
MouseMetaData.wrkin <- MouseMetaData.wrkin[!(is.na(MouseMetaData.wrkin$mouse)|MouseMetaData.wrkin$mouse==""),]
#do this for the mouse table instead of the MLH1 table
MLH1.imprv <- merge(MLH1_data, MouseMetaData.wrkin)#don't apply the all parameters
#DOB is date
MLH1.imprv <- add_euth_date(MLH1.imprv)
#good euth date..

#make backup date col #add euth date for calculating age
MLH1.imprv$euth.date2 <- MLH1.imprv$euth.date
MLH1.imprv <- add_age(MLH1.imprv)
#age works

#remove mice that are older than 20 weeks
MLH1.imprv.male <- MLH1.imprv[MLH1.imprv$sex == "male", ]
MLH1.imprv.male <- MLH1.imprv.male[MLH1.imprv.male$age.weeks < 20, ]

age.hist <- ggplot(MLH1.imprv[MLH1.imprv$sex == "male", ], aes(age.weeks))+geom_histogram()+facet_wrap(~category)+ggtitle("Male Age and MLH1 counts by subspecies")
age.hist

Make a subsetted dataset for plotting mean mouse (quality < 5) MLH1 count and age.

```{r age.effect.model, include=FALSE}
#run a age effect linear regression!

male.data <- AP_mouse_table_w.Ages[AP_mouse_table_w.Ages$sex.y == "male",]

male.data.dom <- male.data[male.data$subsp.y == "Dom",]

male.data.musc.low <- male.data[(male.data$strain.x == "KAZ" | male.data$strain.x == "TOM" | male.data$strain.x == "AST"),]

male.data.musc.high <- male.data[(male.data$strain.x == "PWD" | male.data$strain.x == "MSM" | male.data$strain.x == "MOLF" | male.data$strain.x == "SKIVE"),]

male.data.outgroup <- male.data[(male.data$subsp.y == "Spretus" | male.data$subsp.y == "Spic" | male.data$subsp.y == "Caroli"),]

lm.me.dom  <- lm(mean_co ~ age.weeks+strain.x, data = male.data.dom)

lm.me.musc.low  <- lm(mean_co ~ age.weeks+strain.x, data = male.data.musc.low)

lm.me.musc.high  <- lm(mean_co ~ age.weeks+strain.x, data = male.data.musc.high)
#significant differences across strains in musc high

lm.me.OG  <- lm(mean_co ~ age.weeks+strain.x, data = male.data.outgroup)

summary(lm.me.OG)
#for all models, mostly it was that the intercept was significant, and mot much else
```

```{r female.age.effect, echo=FALSE}
#start thinking of the maternal age effect

#female age effect

#female.meta.data <- meta.data[meta.data$sex == "female",]
#how much data to I have to work with?
#there's a fair number of rows without emb or neonate
#I could probably assume all of these are neonates

#length( !is.na(female.meta.data$female.stage) )

#female.meta.data <- female.meta.data[female.meta.data$female.stage]

#how much formating do I have to do?
#can't use AP_mouse_table since that is summarized
#meta.data
#grepl notes, "neo"

#some have e-number for embryonic
#I might have to

>>>>

```{r setup, echo=FALSE, include = FALSE}
library(knitr)
library(ggplot2)
#library(pwr)
library(plyr)
library(lattice)
library(dplyr)
library(raster)
library(lme4)
library(nlme)
library(RLRsim)

setwd("C:/Users/alpeterson7/Documents/MLH1repo/")
#setwd("C:/Users/April/Desktop/MLH1repo")
#load(file="data/MLH1/MLH1_data_setup.RData")

#load main data file

load(file="data/MLH1/MLH1_data_setup_11.12.19.RData")#added batch17
#load(file="data/MLH1/MLH1_data_setup_11.11.19.RData")#deleted DUP image
#load(file="data/MLH1/MLH1_data_setup_9.26.19.RData")#added batch16
#load(file="data/MLH1/MLH1_data_setup_8.29.19.RData") #
#load(file="data/MetaData.RData") #this was replacing most updated data file

@@

```{r organizing.data, echo=FALSE }
#set the order of another column, based on another variable. consistant order for ploting

@@

@@

kable(table(AP_mouse_table.HQ$category), col.names = c("Category", "mice"), label = "Number of mice by category" )

@@

```{r rand, echo=FALSE, include=FALSE, eval=FALSE}

str(AP_mouse_table_w.Ages)

AP_mouse_table.HQ$var <- as.numeric(AP_mouse_table.HQ$var)

#remove sex.x == na, and strain.x=na or F1
AP_mouse_table.HQ <- AP_mouse_table.HQ[!(is.na(AP_mouse_table.HQ$sex) | AP_mouse_table.HQ$sex==""), ]

AP_mouse_table.HQ <- AP_mouse_table.HQ[!(is.na(AP_mouse_table.HQ$var) | AP_mouse_table.HQ$var==""), ]
#plot variance and mean (facet sex)

#AP_mouse_table_w.Ages <- AP_mouse_table_w.Ages[!( AP_mouse_table_w.Ages$strain.x =="F1"), ]

VandMean.mouse <- ggplot(aes(x=mean_co, y = var), data=AP_mouse_table.HQ )+geom_point(aes(color=strain))+facet_wrap(~sex)+ggtitle("mouse MLH1 averages; variance by mean")

cVandMean.mouse <- ggplot(aes(x=mean_co, y = cV), data=AP_mouse_table.HQ )+geom_point(aes(color=strain))+facet_wrap(~sex)+ggtitle("mouse MLH1 averages; cV by means")

str(AP_mouse_table.HQ)

cor(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "female"], AP_mouse_table.HQ$var[AP_mouse_table.HQ$sex == "female"])
#[1] 0.5750399
cor(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "male"], AP_mouse_table.HQ$var[AP_mouse_table.HQ$sex == "male"])
#0.321601

cor(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "female"], AP_mouse_table.HQ$cV[AP_mouse_table.HQ$sex == "female"])
# 0.3515544
cor(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "male"], AP_mouse_table.HQ$cV[AP_mouse_table.HQ$sex == "male"])
#-0.03713536

cor.test(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "male"], 
    AP_mouse_table.HQ$var[AP_mouse_table.HQ$sex == "male"] )
# p-value = 8.623e-09

cor.test(AP_mouse_table.HQ$mean_co[AP_mouse_table.HQ$sex == "male"], 
    AP_mouse_table.HQ$cV[AP_mouse_table.HQ$sex == "male"] )

#remove mice with fewer than 10 cells

AP_mouse_table.HQ.10plus <- AP_mouse_table.HQ[AP_mouse_table.HQ$Ncells >= 10,]

VandMean.mouse.10plus <- ggplot(aes(x=mean_co, y = var), data=AP_mouse_table.HQ.10plus )+geom_point(aes(color=strain))+facet_wrap(~sex)+ggtitle("10 plus mouse MLH1 averages; variance by mean")

cor(AP_mouse_table.HQ.10plus$mean_co[AP_mouse_table.HQ.10plus$sex == "female"], AP_mouse_table.HQ.10plus$cV[AP_mouse_table.HQ.10plus$sex == "female"])

cor.test(AP_mouse_table.HQ.10plus$mean_co[AP_mouse_table.HQ.10plus$sex == "female"], 
    AP_mouse_table.HQ.10plus$var[AP_mouse_table.HQ.10plus$sex == "female"] )

@@

```{r MLH1.histogram.2421, echo=FALSE, fig.caption= "histograms", warning=FALSE, fig.align='center'}

histM <- ggplot(data=MLH1_data,(aes(x=adj_nMLH1.foci))) + xlim(c(16, 38)) + geom_histogram(stat = "bin", binwidth = 1) + facet_wrap(~ category)+ggtitle("Pooled Cell Histograms")

#when I try to add the mean lines, error returned that it can't find the adj_MLH1 ....?
#histM <- histM +  geom_line(data=mean.MLH1.obs, aes(y=mean.MLH1.obs$mean.MLH1))

#ad
#  geom_smooth(method = lm, se = FALSE
 # geom_smooth(method = lm, formula = mean.MLH1.obs$mean.MLH1 ~ mean.MLH1.obs$category, se = FALSE)

#playing with expected posisson plot...
#function = glm(y ~ x, family = poisson)
#add Poission curve behind
#geom_smooth(method = lm, formula = y ~ splines::bs(x, 3), se = FALSE)
#not sure how to add Expected poisson curve

@@

Most histograms of pooled MLH1 counts are all close to normal. Except PWD male, LEW female, SKIVE male, SPIC male.
check if there is a quality effect.

#  Results
